Search enrichment by protein

### 2025-02-07 Background/MultifunctionalLipidProbesOverview.qmd
EOD push - updates to overview pages, small changes to protein search

## Photoactivatable & Clickable Probes / Bifunctional Lipid Probes

<!-- -   The conceptual precursor to the so-called trifunctional probes, these typically include only the diazirine and the terminal alkyne moieties.

@@

-   [@haberkantFatFabulousBifunctional2014] - Haberkant et al. 2014 -->

## Trifunctional Lipid Probes

@@

* While significant effort has minimized the effect that each modification has on the biophysical properties of the lipid analogs, it remains the case that adding a diazirine and a terminal alkyne will necessarily affect their physical properties -- perhaps affecting sub-cellular localization and binding. 
* The coumarin cage has been repeatedly shown to protect lipid analogs from premature metabolism, but it should be recognized that this comes with some distinct changes to the behavior of the native lipid. 
  - For example, when a cell is treated with a photocaged lipid probe, every membrane in the cell appears to

## Interactomics Experimental schema

::: border
![](/Background/BackgroundFigures/BaselineProteomics_ExperimentalDesign.png){group="Overview_gallery"}
:::

### Cell treatment

### Protein collection

### Mass spectrometry analysis

:::{.callout-note collapse="true"}
#### Label-Free Quantification

Label-free quantification (LFQ) is a widely used mass spectrometry-based proteomics approach that estimates protein abundance without chemical labeling. Instead, it relies on comparing the summed intensity of detected peptides across different samples analyzed under identical conditions.

##### How It Works

1.  **Peptide Detection (MS1):** Peptides are isolated and detected based on their mass-to-charge ratio (m/z).

2.  **Peptide Sequencing (MS2):** Selected peptide ions are fragmented, and the resulting spectra determine peptide sequences.

3.  **Quantification:** The total signal intensity of all identified peptides corresponding to a protein is summed as a proxy for its abundance.

4.  **Comparison Across Samples:** Samples prepared and analyzed using the same LC-MS/MS protocol can be compared using peptide-spectrum match (PSM) counts as an abundance proxy.

##### Advantages of Label-Free Quantification

-   **Cost-Effective & Simple Sample Preparation:** Unlike labeling methods (e.g., TMT), LFQ does not require expensive reagents or complex modifications, making it more accessible.

-   **Flexible & Widely Used:** Compatible with various experimental designs and allows for comparisons across many samples without multiplexing constraints.

##### Challenges & Considerations

-   **Potential Bias from Co-Elution:** High-abundance peptides can overshadow low-abundance ones, reducing the number of reliably quantified peptides.

-   **Run-to-Run Variability:** LFQ requires separate MS runs for each sample, making it susceptible to differences in instrument calibration, chromatography conditions, or sample handling. Careful quality control is necessary to minimize artificial differences.

##### Conclusion

Label-free quantification is a practical and widely used method for proteomics research. While cost-effective and flexible, its accuracy depends on careful experimental design and quality control to reduce variability.
:::
:::{.callout-note collapse="true"}
#### Quantitative Mass Spectrometry with Tandem Mass Tagging {#sec-TMT}

Tandem Mass Tag (TMT) labeling is a powerful quantitative mass spectrometry technique that enables the simultaneous analysis of multiple samples in a single MS run. This approach improves efficiency and allows for precise relative quantification of proteins across experimental conditions.

##### How It Works

1.  **Chemical Labeling:** Each sample is chemically tagged with a unique TMT label, which consists of a reactive group binding to peptides and a mass reporter for sample differentiation.

2.  **Sample Pooling:** Labeled samples are combined into a single pooled group, reducing the need for separate MS runs.

3.  **Mass Spectrometry Analysis:**

    -   **MS1 (Peptide Isolation):** Peptides are separated and detected based on their mass-to-charge ratio (m/z).

    -   **MS2 (Peptide Sequencing & Tag Fragmentation):** Peptides are fragmented for sequence identification, and TMT mass tags fragment to generate reporter ions.

4.  **Quantification:** The intensity of reporter ions reveals the relative abundance of peptides across the original samples.

5.  **Comparison Across Samples:** Within a TMT cassette (typically 16 samples), relative protein abundance is calculated from the reporter ion intensities. A pooled control channel enables comparison across cassettes, allowing up to 30 samples to be analyzed in a single experiment.

##### Advantages of TMT Labeling

-   **Increased Throughput & Efficiency:** Enables multiplexing of many samples (e.g., 16 or more), reducing the number of separate MS runs.

-   **Deeper Proteome Coverage:** Pooled samples can be fractionated (e.g., by cation exchange chromatography), enhancing detection of low-abundance peptides.

-   **Improved Quantification Accuracy:** Since all samples are analyzed in the same MS run, technical variability between runs is minimized.

##### Challenges & Considerations

-   **Higher Cost:** TMT reagents and high-end mass spectrometers add to the expense.

-   **Complex Sample Preparation:** Requires additional steps for labeling, pooling, and fractionation, increasing the risk of sample handling variability.

-   **Normalization Requirement:** Accurate quantification requires equal protein input across samples to avoid artificial balancing biases.

-   **Demanding Instrumentation & Expertise:** Requires high-resolution mass spectrometers, often necessitating specialized facilities.

##### Conclusion

TMT-based mass spectrometry is a valuable tool for large-scale proteomics, offering high-throughput, precise, and deep proteome coverage. While it requires advanced instrumentation and careful sample preparation, its ability to multiplex multiple samples in a single run makes it an attractive option for high-confidence protein quantification.
:::

>>>>

<!--
## Photoactivatable & Clickable Probes / Bifunctional Lipid Probes

 -   The conceptual precursor to the so-called trifunctional probes, these typically include only the diazirine and the terminal alkyne moieties.

@@

-   [@haberkantFatFabulousBifunctional2014] - Haberkant et al. 2014

## Trifunctional Lipid Probes
-->

@@

* While significant effort has minimized the effect that each modification has on the biophysical properties of the lipid analogs, it remains the case that adding a diazirine and a terminal alkyne will necessarily affect their physical properties -- perhaps affecting sub-cellular localization and binding. 
* The coumarin cage has been repeatedly shown to protect lipid analogs from premature metabolism, but it should be recognized that this comes with some distinct changes to the behavior of the native lipid. 
  - For example, when a cell is treated with a photocaged lipid probe, the probe intercalates into every membrane in the cell -- including membranes which the native lipid would not enter. This may lead to the identification of non-biologically-relevant hits. 
* Similarly, the diazirine and alkyne groups may have minor effects on the biophysical properties of the lipid probe as it interacts with other lipids and proteins within the membrane bilayer. This may be more relevant for proteins which preferentially bind to the acyl chain. 
  - For example, Some work has been done to demonstrate that the placement of the diazirine ring along the length of the acyl chain can have impacts on the interactome of the lipid probe[@farleyTrifunctionalFattyAcid2024]. Farley et al (2024) demonstrated this using trifunctional Fatty Acid probes by placing the diazirine either 3 carbons from the terminus (8-3 FA) or 10 carbons from the terminus (1-10 FA). They identified several proteins with distinct enrichment profiles, including VDAC1 and VDAC2 (which are preferentially enriched to the 8-3 FA probe), and ACAD9 and UBB (which are preferentially enriched to the 1-10 FA probe). 
* Quantification of +UV versus -UV
  - The details of the quantitative mass spectrometry are describe in our [Overview of Proteomics Methods](Background/ProteomicsUsingMultifunctionalProbes.qmd), however, we hope to succinctly describe that quantitative mass spectrometry techniques necessarily require normalization at sample input level -- thereby exaggerating the abundance of proteins in the -UV control samples. 
  - It should be noted, therefore, that there is little meaning to a protein which is "enriched" toward the -UV condition. For this reason, all proteins which are "depleted" following irradiation with UV have been designated as "no hit" in our analyses.

### 2025-02-07 Background/ProteomicsUsingMultifunctionalProbes.qmd
EOD push - updates to overview pages, small changes to protein search

- Irradiation with 355 nm wavelength light ()

@@

- Cell lysis

- Affinity purification of probe-conjugated proteins
  * Example using azide beads: Perform click chemistry reaction in presence of beads coated with picolyl-azide functional groups for covalent capture
  * Conjugates of protein-lipid-bead are covalently bound and can thus be stringently washed with detergent solutions to remove non-specific proteins. Detergent can then be removed with repeated washes of 8 M urea and 20 % acetonitrile. 
* Captured proteins can be reduced and alkylated on-bead. 
* Peptides can be digested off the protein-lipid-bead conjugates via tryptic digest. Digested peptides are then desalted to remove any remaining contaminants using reverse phase desalting columns.

### Mass spectrometry analysis

>>>>

- Irradiation with 355 nm wavelength light induces the production of a reactive carbene which readily inserts itself into any nearby biochemical bond. 
  - Notably, the majority of probes will likely crosslink with either water or with neighboring lipids. Quantitative proteomics analysis will fully ignore these conjugates, as only enriched peptides will be measured in the final mass spectrometry analysis (described below in [@sec-MassSpectrometryAnalysis]). 
- The covalently bound lipid protein conjugates which form after crosslinking are highly stable and thus are ideally amenable toward stringent enrichment and washing, producing a high-quality sample with few non-specific proteins tagging along.

@@

- Cell lysis 
  - Depending on application, may want to perform a membrane prep or otherwise isolate an organelle of interest (e.g. mitochondrial enrichment, if one is seeking mitochondrial interactions)
  - Due to the high stability of the diazirine crosslinking bond, no extra steps need to be taken to ensure that interactions are maintained -- lysis can be as aggressive as one wishes.
  - In general, samples due for mass spectrometry analysis should be kept with as little detergent as possible to protect the instrument -- however, because the crosslinking bond is covalent and the downstream affinity purification is covalent, these samples may be washed very stringently and thus remove effectively all contaminating detergents. Thus, a mild detergent such as NP-40 can be included in the lysis buffer of these samples.

- Affinity purification of probe-conjugated proteins
  * Example using azide beads: Perform click chemistry reaction in presence of beads coated with picolyl-azide functional groups for covalent capture
  * Conjugates of protein-lipid-bead are covalently bound and can thus be stringently washed with detergent solutions to remove non-specific proteins. Detergent can then be removed with repeated high volume washes of 8 M urea and 20 % acetonitrile. 
* Captured proteins are then reduced and alkylated on-bead. 
* Peptides are digested off the protein-lipid-bead conjugates via tryptic digest. Digested peptides are then desalted to remove any remaining contaminants using reverse phase desalting columns.

### Mass spectrometry analysis {#sec-MassSpectrometryAnalysis}

### 2025-02-07 SearchByProtein/SearchByProtein.qmd
EOD push - updates to overview pages, small changes to protein search

## Displays the reactive dataframe
datatable(datawhole_reactive,
          caption = "'No hit' proteins are unenriched or depleted in the presence of probe, 'enriched candidates' are defined as proteins with a false discovery rate less than 0.2 and a fold change of at least 1.5-fold, and 'enriched hits' are defined as proteins with a false discovery rate less than 0.05 and a fold change of at least 2-fold in the +UV over the -UV", 
          options = list(pageLength = 6,
												 dom = "l<t>ipr"))

@@

>>>>

## Displays the reactive dataframe
datatable(datawhole_reactive, 
          options = list(pageLength = 6,
												 dom = "l<t>ipr"),
												 colnames = c("Gene Name", "Lipid Probe", "Hit Annotation", "logFC", "pvalue", "FDR", "Study Details"))

@@

"No hit" proteins are unenriched or depleted in the presence of probe, "enriched candidates" are defined as proteins with a false discovery rate less than 0.2 and a fold change of at least 1.5-fold, and "enriched hits" are defined as proteins with a false discovery rate less than 0.05 and a fold change of at least 2-fold in the +UV over the -UV
